---
title: "A Gentle Introduction to R"
author: "Victor Ordu"
date: "6 August 2018"
output: ioslides_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Outline

* Data
* Functions
* The Working Directory
* The Workspace
* R Script
* Resources

***

# Data
***

Everything about computing has to do with data - R is no exception.

* The user provides input.
* The input is processed.
* R gives either gives an output or carries out an action e.g. open a file.

```{r hello}
# This is a comment. R ignores this line.

# Type in "Hello, world!"
"Hello, world!"
```

```{r num}
42
```

***

R can perform computations
```{r}
18 + 4
```

```{r subtract}
22 - 4
```

```{r}
18 * 6
```

```{r}
108 / 6
```

***

```{r}
18 %% 5
```

```{r}
3 ^ 2
```

## Variables

* In order to work effectively with pieces of data, we give them identifiers known as `name`s or `symbol`s.
* R offers a special operator called the **assignment operator** and represented by the synbol *`<-`*
* The operator `=` can also be used but it's not recommended.

* The assignment operator is used to connect a value to a name.

So, going by one of the earlier examples

```{r, vars}
x <- 18
```
We now have a piece of data in memory called `x ` that has a value of 18.

***

If one runs `x` in the console, any value it has should be returned as output.

```{r var output}
x
```

It's that simple!

* We now have a variable `x` stored in memory.
* It is called a variable because we can change the value but retain the name.

```{r change var}
x <- 108
x
```

Voila! Our variable `x` has just been assigned a new value.

***

We can add create another variable and use them for computations

```{r}
y <- 6
x + y
```

Okay, let's put the results into a third variable.

```{r}
z <- x + y
z
```

*Quiz*: What happened here?

***

This manipulation of variables-values is essentially what we do in any R session, whether on a small or large scale, so need to understand this concept.

One common trick is to use an existing variable to alter its own value.

```{r}
x
x <- x / 12
x
```

# Functions
***

Every thing that happens in R is done with a function.

For example we have a function in R for computing the square root of a number called `sqrt`

```{r}
x
sqrt(x)
```

What is the value of `x` now?

***

R functions are usually represented in the format `<name>` followed by  `<parentheses>` i.e. `()`.

The parentheses are used to "wrap" the function parameters. Note that not all functions have parameters.

The user can also create his own functions

## The Working Directory
When working in R, you are always within a given working directory. To know the current working directory call the function `getwd`.

```{r wd}
getwd()
```

To change the working directory, write the path to the new directory you want to move to. For example

```{r, eval=FALSE}
setwd("~/Desktop")
```

To check where you are now, run `getwd()` again.

***

It is strongly recommended that for any given project, you have a unique working directory, hence the introduction of this function.

# The Workspace

***

This is essentially your R workstation and this is where all the data you are working with in any given session resides. To display the contents of the workspace, use the `ls()` function.

```{r}
ls()
```

***

To remove something from the Workspace, use the `rm()` function.

```{r}
rm(z)

```

To remove *everything* in the Workspace supply additional arguments to `rm`. 

* This function has a parameter `list`
* We will supply the argument that returns everything in the workspace i.e. `ls()`.

```{r}
rm(list = ls())
```

# R Scripts

***

* R commands can be saved in a text file called an R script.

* R reads the file line by line from top to bottom.

* Each line is read from right to left.

* The implication of this is that when there is a problem with the code, R stops execution at the line with the syntax issue. Errors on subsequent lines will only be read after it is fixed.

* Similarly, if there is more than one problem on a given line, the right-most is read first and exexution stops at that point. 

* R does not foresee errors beyond the point where execution is stopped

# Code Demo

***

# Resources
## R Documentation
* Your R installation comes with a rich set of documentation that can be easily accessed via certain functions.
* This includes the R Manuals, such as language definitions, general information, data importa/export details, installation tips and even the nuts-and-bolts of the R interpreter.
* To see the manuals, you can call `help.start()` to open the main Help homepage.

## Help
* Specific help can be easily obtained for indidual functions, amongst others.
* For example to see the help of `sqrt()` function, all we need to is to prepend the name of the function with the operator `?`.
* This will open a function documentation page with details on usage, arguments, return value and perhaps examples.

***

Sometimes, you just want to check the arguments to use
```{r}
args(setwd)
```

Or you just want to see some example(s)
```{r}
example("sum")
```

# Thank you!