---
title: "Introducing NARC Data Projects"
author: "Victor Ordu"
date: "23-24 August 2018"
output: ioslides_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = FALSE)
```

## Introduction
* Earlier sessions ->> equip to **use** R.
    + Can navigate an R session
    + Can create simple R objects
    + Can examine R objects
    + Can carry out basic object manipulations
    + Can look for help!

## Problem Statement
Our findings:
<ol>
<li> The data existed in different Microsoft Excel files, some of which had more than one spreadsheet. </li>
<li> The data in each spreadsheet had different data fields </li>
<li> The principal data collection instrument was the Guest Form of the ministry. </li>
<li> The newsletter mailing list was being managed with MailChimp, a popular online mailing list management platform. </li>
<li> The data were being manually uploaded to MailChimp, specifically the names and email addresses of prospective readers. </li>
</ol>
***
<ol start=6>
<li> The staff handling the data did not have sufficient knowledge of simple data management practices. </li>
<li> There is no documented data management policy at the Centre and thus the staff do not have a specific workflow that they employ and as such data entry methods are arbitrary. </li>
<li> There is no in-house mechanism for monitoring data quality. </li>
<li> There is no back-up for the data. </li>
</ol>

## Solution
The solution for these problems are domiciled in:

1. An R package called _**NARCcontacts**_.
2. An RStudio project called _**narc-mail-data**_.

Both of them can be found via the DevSolutions GitHub page.

***

**NARCcontacts**

1. Contains functions that specifically address some of the problems identified.
2. Makes it easy for the staff to run the operations
3. Provides avenue for improvement and fixing unforeseen issues
4. The approach used can be reproduced and shared - this is good for learning.

***

**narc-mail-data**

* This project is a folder that contains everything associated with the intervention.
    + Most notable is the data - original and modified.
    + Reports
    + Further development and experimentation i.e. sandbox.
* Data entry will in the interim be carried out here as well.

# Usage
## Git
* Now that we have **git** installed, let's make sure. 
* Open a shell of your choice
* At the prompt, run the command `git`
You should see something like this:

```
usage: git [--version] [--help] [-C <path>] [-c name=value]
           [--exec-path[=<path>]] [--html-path] [--man-path] [--info-path]
           [-p | --paginate | --no-pager] [--no-replace-objects] [--bare]
           [--git-dir=<path>] [--work-tree=<path>] [--namespace=<name>]
           <command> [<args>]

These are common Git commands used in various situations:

...
```
***
To collect a copy of the R Studio project **narc-mail-data**:

* open the repository main page (I will show you how!) and copy the URL from the address bar.
* run this command in the shell
```
git clone <full web address>.git
```
* Note that you're to `.git` at the end.
* This will give you the whole project folder.

## R packages
Now you will install the R package.  

* Because your R installations are relatively new, **do not start the package installation until we are through with the talk**!
* Write out the package installation code in an R script while we discuss.

***

To use an R package, there are 2 main steps:

1. Download and installation
2. Attaching and/or loading

* The first brings the package onto your machine
* The second bring the package into your current R session

***

* The main package repository for R is called the *Comprehensive R Archive Network* (CRAN), owned by the R Foundation for Statistical Computing (see <http://r-project-org)

* Another important (unofficial) repository is GitHub - this is where most development versions of packages are found.

* Yet another is called **Bioconductor**.

Anyone can create packages and host online for others to use.

***
* To install from CRAN, we use the function, `install.packages()`. You are going to need a package called `devtools` to easily install from GitHub.
```{r}
install.packages('devtools')
```

* To install a package from *GitHub* we use `devtools::install_github()`. To install our *NARCcontacts*, use this code
```{r}
devtools::install_github('DevSolutionsLtd/NARCcontacts')
```

# Project Overview
## Contact List Harmonisation
* To harmonise the data that exists in different Excel files
* The function that does this is `NARCcontacts::harmonise_narc_excel()`.
* Its arguments are
    + `dir` - A directory that contains the Excel files to be merged.
    + `dest` - The folder where the resultant database is to be stored. Defaults to the **data** folder in *narc-mail-data*. 
    + `quiet` - Whether to display output on the progress of the operations. Defaults to `FALSE`.
    
***

* Can be activated by `source('src/harmonise.R')`
* Others: *Source* button, PowerShell script

Example:
```{r}
harmonise_narc_excel("workbooks", "myFolder")
```

## Data Consolidation
* Upon merging the files, invariably some of the records were repeated.
* Many of the duplicated records were inconsistent.
* `NARCcontacts::consolidate_narc_mail()` fixes this.
* Fixes are applied interactively.
* Function receives 1 argument, `db`
* Currently ends with a warning - to be fixed later.
* Can be activated in *narc-mail-data* by `source('src/consolidate.R')`

## Data Entry
* The *narc-mail-data* project also has a script for activating a data entry form that is based on the __DataEntry__ package.
* To open the form simply type `DataEntry::DataEntry()` or source the script **dataentry.R**. 

## Bonus: Media Management
* **Purpose**: To find all the media files in a given system.
* Run (source) the script `locate-msg.R`.
* The user will be prompted with a dialog to select a root folder to search for media files.
* An entire drive (e.g. `C:/` on Windows) can be searched this way. If you do this, go and make a cup of coffee!
* Details of the media files will be stored in a database - useful for later searches.

## Future Work
See `README`

# Live Demo

# Conclusion
## Next Steps
* Git & GitHub Tutorial
* Basic (project specific) Data Analysis 
* Basic R Programming

**But from here on everything will be pretty much hands-on!**

## Contact
Feel free to contact me via email and follow me on Twitter & GitHub  

**Email**: victor@dev-solu.com  
**Twitter**: [BroVic](https://twitter.com/BroVic)  
**GitHub**: <https://github.com/BroVic>  